VariantSpark: Population Scale Clustering of Genotype Information

### 2015-11-26 doc/publications/journal2014/paper.tex
updated paper and letter

%To further demonstrate the scalability of \variantSpark{}, we also cluster the larger 1000 Genomes Project phase 3 VCF files. Although \kMeans{} clustering completes successfully on this dataset,
%it takes over 27 hours (see Table 2) due to more than twice the number of individuals, and hence more variants. For comparison, the uncompressed size of the phase 3 files is 770GB vs the 161GB of the phase 1 dataset.
%TODO add value
To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 superpopulations and as a result has over 80 Million variants. 
The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
\variantSpark{} successfully completes the clustering in just 27 hours (see Table 2) with an ARI of XX. % if XX is smaller than 0.84 we need to explain more.

@@

%\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 million variants in under 15 hours using minimal memory (24GB). 
\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 27 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
%TODO added sentence
Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g. identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g. capture medical history as well as predispositions for diagnosis and treatment decisions.

@@

\subsection*{R Implementation}
For our R implementation, we read in a VCF file using {\sc read.table}, and then convert the resulting table to a matrix ({\sc vcfMatrix} with {\sc as.data.frame.matrix}.
%TODO VariantAnnotation reads in the file quicker but causes issue down the track so overall read.table is faster, right ? - can you add this explanation to the supplementary material.
In the supplementary material we benchmark this approach against reading the VCF file in using the `VariantAnnotation' package.
We remove the first 9 columns which leaves us with just the variants. As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
We then transpose the matrix with {\sc t(vcfMatrix)}, which results in a data-structure where each row represents an individual. We convert the matrix to a {\sc big.matrix} object,
as required by the \kMeans{} algorithm from the `biganalytics' package (\url{https://cran.r-project.org/web/packages/biganalytics/index.html}), and then call {\sc bigkmeans} with
the {\sc big.matrix} object and the required number of clusters as arguments.

>>>>

%To further demonstrate the scalability of \variantSpark{}, we also cluster the larger 1000 Genomes Project phase 3 VCF files. Although \kMeans{} clustering completes successfully on this dataset,
%it takes over 27 hours (see Table 2) due to more than twice the number of individuals, and hence more variants. For comparison, the uncompressed size of the phase 3 files is 770GB vs the 161GB of the phase 1 dataset.
To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 super-populations and as a result has over 80 Million variants. 
The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
\variantSpark{} successfully completes the clustering in just 27 hours (see Table 2) with an ARI of 0.82.

@@

%\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 million variants in under 15 hours using minimal memory (24GB). 
\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 27 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g. identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g. capture medical history as well as predispositions for diagnosis and treatment decisions.

@@

\subsection*{R Implementation}
For our R implementation, we read in a VCF file using {\sc read.table}, and then convert the resulting table to a matrix ({\sc vcfMatrix} with {\sc as.data.frame.matrix}.
%TODO VariantAnnotation reads in the file quicker but causes issue down the track so overall read.table is faster, right ? - can you add this explanation to the supplementary material.
In the supplementary material we benchmark this faster approach against reading the VCF file in using the `VariantAnnotation' package.
We remove the first 9 columns which leaves us with just the variants. As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
We then transpose the matrix with {\sc t(vcfMatrix)}, which results in a data-structure where each row represents an individual. We convert the matrix to a {\sc big.matrix} object,
as required by the \kMeans{} algorithm from the `biganalytics' package (\url{https://cran.r-project.org/web/packages/biganalytics/index.html}), and then call {\sc bigkmeans} with
the {\sc big.matrix} object and the required number of clusters as arguments.
